\documentclass[12pt]{article}
\usepackage{amsmath}
%\usepackage[showframe]{geometry}
\setlength{\voffset}{-0.75in}
\addtolength{\textheight}{1in}
%\title{Test}
\date{}
\pdfpagewidth 8.5in
\pdfpageheight 11in
\pagestyle{empty}
\begin{document}%
	\begin{center}
	\textbf{\large{Cole Nielsen}}
	
	4748 Hummingbird Trail
	
	Prior Lake, MN 55372
	
	niels538@umn.edu
	
	952-607-8428
	\end{center}
	\textbf{Education}
	\newline
	{University of Minnesota - Minneapolis, MN (Fall 2014 - present)}
	\newline\indent\textit{{B.S. - Electrical Engineering, expected graduation May 2017}}
	\newline
	{University of Minnesota - Minneapolis, MN (Fall 2013 - Spring 2014)}
	\newline{\indent{\textit{Full time PSEO student, Deans List Fall Semester, GPA 3.65}}}
	\newline
	{Prior Lake High School - Savage, MN (Fall 2010 - Spring 2014)}
	\newline{\indent{\textit{Diploma, Board Scholar, GPA 3.889}}}
	\newline\newline
	\textbf{Experience}
	\newline{Circuitcell (6/2014 - present)}
	\newline\indent{\textit{Designed and built battery test fixture, assembled product.}}
	\newline{Mystic Lake (8/2012 - present)}
	\newline\indent{\textit{Hosted and bussed. Customer service experience.}}
	\newline{First Robotics - FRC (2010 - 2014)}
	\newline\indent{\textit{Captain and lead of robot mechanical design during 2014 season for team 2169 KING TeC. Used and gained 		proficiency with 3D CAD software (Autodesk Inventor), mechanical drawings, assembly and design.}}
	\newline{First Robotics - FTC (2009 - 2010)}
	\newline\indent{\textit{Mechanical and electrical design/assembly of team robot.}}
	\newline\newline
	\textbf{ Recent Personal Projects}
	\newline{Tesla Coils:} \textit{Completely built and designed several solid state tesla coils from}
	\newline\indent{\textit{scratch, including the PCBs (designed in Eagle PCB, manually etched).}}
	\newline{Amplifier: } \textit{Built and designed a 2.1 channel amplifier including the mechani-}
	\newline\indent{\textit{cal assembly, circuitry and PCBs.}}
	\newline{Repaired broken Ludlum radiation survey meter:} \textit{Reverse engineered and}
	\newline\indent{\textit{then diagnosed the issues with the meter and fixed them.}}
	\newline\newline
	\textbf{Skills}
	\newline{Electronics: \textit{EDA software (Eagle PCB, KiCAD), SPICE, Soldering, circuit design, microcontrollers, know how to operate test equipment like oscilloscopes, function generators etc.}}
	\newline{Mechanical design: \textit{3D CAD software (Good with Inventor, can use Solidworks)}}
	\newline{Software and programming: \textit{Linux and Windows, Microsoft office, C/C++ (improving skills with C/C++ class now), EDA/CAD.}}
	
\end{document}